---
title: Z(log) Transformation for Laboratory Measurements
output:
    rmarkdown::html_vignette:
        toc_float: true
vignette: >
    %\VignetteIndexEntry{Z(log) Transformation for Laboratory Measurements}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
bibliography: bibliography.bib
---

# Introduction

The `zlog` package offers functions to transform laboratory measurements into
standardised *z* or *z(log)*-values as suggested in [@hoffmann2017].
Therefore the lower and upper reference limits are needed. If these are not
known they could estimated from a given sample.

# Z or Z(log) Transformation

Example data and reference limits are taken from [@hoffmann2017, Table 2].

```{r zlog}
library("zlog")

albumin <- c(42, 34, 38, 43, 50, 42, 27, 31, 24)
z(albumin, limits = c(35, 52))
zlog(albumin, limits = c(35, 52))
```

# Estimate Reference Limits

The `reference_limits` functions calculates the lower and upper 2.5 or 97.5
(or a user given probability) quantiles:

```{r reference_limits}
reference_limits(albumin)
reference_limits(albumin, probs = c(0.05, 0.95))

exp(reference_limits(log(albumin)))
```

# Session information

```{r si}
sessionInfo()
```

# References
